One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2024-10-31 paper.tex
edits up to results section

\vfill

@@

To motivate the approach, we consider a linearised approximation of $\eta(\bm{u})$ around point $\bm{u}_0$:
\begin{equation*}
\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0),
\end{equation*}
where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.  Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the first-order Taylor approximation $\bar{\eta}_{\bm{u_0}}$.  This model configuration has $\bm{B}_{\bm{u}_0}$ as the model design matrix and $\eta(\bm{u}_0} - \bm{B}_{\bm{u}_0}\bm{u}_0$ as a known offset term.

@@

The `optimal' approximate model is defined as a fixed point of this functional.  Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_{\bm{v}})$ to use as the new linearisation point.  We iterate this procedure until a fixed point $\bm{v}^*$ is identified within a chosen tolerance such that $\bm{v}^* = f(\bar{p}_{\bm{v}^*})$.

\section{Spatially Structured Random Effect}
\label{sec-gmrf}

This section describes the spatial model for the intensity process.  
As mentioned above, the \akepa{} population exhibits spatial heterogeneity for which there is no known explanatory covariate.  Therefore we use a spatially structured random effect to explain this variation.

Using spatially-structured random effects can be computationally expensive.  Often this is due to parameterisations and approximations that lead to dense prior precision matrices which have an  $\mathcal{O}(k^3)$ cost to factorise for an effect with $k$ basis parameters.  A key approach in computationally efficient methods for spatially structured random effects is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).  These result in a sparse precision matrix which can be factorised with a cost of $\mathcal{O}(k^{3/2})$.  We use a GMRF representation based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.  This representation is constructed by applying finite element methods to solve the SPDE numerically.

Since only the distance to the observer was recorded, and not the exact location of \akepa{}, we used the adjusted intensity given in equation \eqref{intensity-incomplete}.  The log-intensity of animal locations, detected or undetected, is given by
\begin{equation*}
\log \lambda(\bs) = \beta_0 + \xi(\bs)
\end{equation*}
where $\beta_0$ is an intercept parameter and $\xi(\bs)$ is a mean zero GRF with Mat\'ern covariance.  The Mat\'ern covariance function in two dimensions can be written as 
\begin{equation}
C(\bs_1,\bs_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|\bs_1-\bs_2\|)^{\nu}K_\nu(\kappa \|\bs_1-\bs_2\|)
\end{equation}
where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.  The three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which specifies the mean-square differentiability of the process.  We set $\nu = 1$ which is the default value in \texttt{R-INLA}.

We use a GMRF representation of $\xi(\bs)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(\bs)$ via a SPDE.  We define a finite element mesh with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$. The GRF is represented as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.  The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (\cite{blangiardo_spatial_2013}, Equation 6.19). The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function as distance between locations increases. We use a finite element mesh with 3748 nodes defined over the entire study region and extended over a buffer region to avoid the effects of boundary conditions within the region of inference.

In order to specify priors on the Mat\'ern covariance we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.  When $\nu = 1$ and the domain is two dimensional then the reparameterisation is $\rho = \sqrt{8} / \kappa$ for the range and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ for the marginal variance (\cite{blangiardo_spatial_2013}, p196).  This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.  For the Mat\'ern field, the base model is the limiting case with $\sigma^2 \rightarrow 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.  The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be essentially `turned off' if the model component is not supported by the data.  To specify penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.  The value of 130 for the range parameter was selected based on the minimum distance between sampling locations.

The SPDE approach can at first appear to be fundamentally different  other approaches such as penalised regression splines.  However, the two methods are closely related and the GRF model component in our model plays a similar role to a penalised smoothing spline on space.  \cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs and \cite{miller_UnderstandingStochasticPartial_2020} do the reverse to show how to implement the Mat\'ern SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.

>>>>

@@

To motivate the approach, we consider a linearised approximation of $\eta(\bm{u})$ around point $\bm{u}_0$:
\begin{equation*}
\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0),
\end{equation*}
where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.  Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the first-order Taylor approximation $\bar{\eta}_{\bm{u_0}}$.  This approximate model configuration has design matrix $\bm{B}_{\bm{u}_0}$ and known offset term $\eta(\bm{u}_0) - \bm{B}_{\bm{u}_0}\bm{u}_0$.

@@

The final linearisation point is defined as the fixed point of this functional.  Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_{\bm{v}})$ to use as the new linearisation point.  We iterate this procedure until a fixed point $\bm{v}^*$ is identified within a chosen tolerance such that $\bm{v}^* = f(\bar{p}_{\bm{v}^*})$.  In other words, we iteratively search for a location in the parameter space for $\bm{u}$ where the Taylor expansion is taken about the posterior mode.

\subsection{The intensity process model}
\label{sec-gmrf}

As mentioned above, the \akepa{} population exhibits spatial heterogeneity for which there is no known explanatory covariate.  To address this we use a sparse Gaussian Markov random field (GMRF) representation of Gaussian random field (GRF) with Mat\'ern covariance based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.

We use the adjusted intensity given in equation \eqref{intensity-incomplete} for incomplete data.  The log-intensity of animal locations, detected or undetected, is given by
\begin{equation*}
\log \lambda(\bs) = \beta_0 + \xi(\bs)
\end{equation*}
where $\beta_0$ is an intercept parameter and $\xi(\bs)$ is a mean zero GRF with Mat\'ern covariance
\begin{equation}
C(\bs_1,\bs_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|\bs_1-\bs_2\|)^{\nu}K_\nu(\kappa \|\bs_1-\bs_2\|),
\end{equation}
where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.  The three covariance function parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and the SPDE approach requires us to assume a value for $\nu$ which specifies the mean-square differentiability of the process.  We set $\nu = 1$ which is the default value in \texttt{R-INLA}.

We define a finite element mesh with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$. The GRF is represented as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.  The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (\cite{blangiardo_spatial_2013}, Equation 6.19). See \cite{lindgren_explicit_2011} for full details.

The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function as distance between locations increases.  In order to specify priors on the Mat\'ern covariance we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.  When $\nu = 1$ and the domain is two dimensional then the reparameterisation is $\rho = \sqrt{8} / \kappa$ for the range and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ for the marginal variance (\cite{blangiardo_spatial_2013}, p196).  This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.  The base model is the limiting case with $\sigma^2 \rightarrow 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.  We set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.  The value of 130 for the range parameter was selected based on the minimum distance between sampling locations.

The SPDE approach can at first appear to be fundamentally different to other spatial effects such as penalised regression splines, which are more typically used in the two-stage approach.  However, these two methods are closely related and the GRF model component in our model plays a similar role to a penalised smoothing spline.  \cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs. \cite{miller_UnderstandingStochasticPartial_2020} present the reverse perspective and show how to implement the Mat\'ern SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
